---
title: "R cheatsheet - data tables"
output: html_document
author: David Sky - david.s.toronto@gmail.com
---
Lets load some sample data into a data table to work with:

```{r}
require("data.table")
sampleDT <- read.table("people-data.csv", header = T, sep=",", stringsAsFactors = F)
```
(Documentation for read.table() at https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.table.html )

## Understanding the data we loaded

Quick commands to better understand the data table:

Operation |R command|Result
----------|---------|------
Count the number of rows | nrow(sampleDT) | `r nrow(sampleDT)`
Count the number of columns |ncol(sampleDT) | `r ncol(sampleDT)`


- use nrow(sampleDT) to determine the number of rows: `r nrow(sampleDT)`
- use ncol(sampleDT) to determine the number of columns: `r ncol(sampleDT)`

Display the column names: 
```{r}
names(sampleDT)
```

Because we included the header = T attribute in the read.table() call, we can use the column names directly.

### References:
- More details on using R Markdown: http://rmarkdown.rstudio.com
- An R Markdown cheat sheet: http://shiny.rstudio.com/articles/rm-cheatsheet.html
