---
title: "R cheatsheet - data tables"
output: 
    html_document:
        toc: true
        toc_depth: 3
        includes:
            after_body: include/data-table-common-hits.html

author: David Sky - david.s.toronto@gmail.com
---
# Introduction
As I work more and more with R I find myself repeating the same tasks, and the most obvious place to start is with loading data into R and better understanding the structure. The following steps are pretty common for working with any data in R. (I authored this [and my other R cheat sheets](index.html) directly in R Markdown[^2] which automatically displays the R code, runs the code, then includes the output in the HTML. )

# Load sample data using the read.table() R function
Lets load some sample data into a data table to work with using the R `read.table()` function [^1]. 

```{r}
require("data.table")
sampleDT <- read.table("people-data.csv", # The people-data.csv file from the current working directory 
                       header = T, # Our CSV file has column header information in the first line
                       sep=",", # Our CSV file uses a comma to separate each column
                       stringsAsFactors = F, # false so that each column is loaded as a set of strings
                       colClasses=c("Followers"="numeric", "Dog.person"="factor")
)
head(sampleDT)
```

# Explore the data we loaded

R provides a lot of ways to quickly get a sense of data - data types, dimensions, ranges of values, etc...

## Quick R commands to understand a data table

Because we included the `header = T` attribute in the `read.table()` call we can use the column names directly in these sorts of commands:

Operation |R command|Result
----------|---------|------
Display the dimensions | `dim( sampleDT )` | `r dim(sampleDT)`
Or just the row count | `nrow( sampleDT )` | `r nrow(sampleDT)`
...and column count | `ncol( sampleDT )` | `r ncol(sampleDT)`
Find the column names | `names( sampleDT )` | `r names(sampleDT)`
Factors in the Dog person column | `levels( sampleDT$Dog.person )` | `r levels(sampleDT$Dog.person)`
Count the number of factors | `length(levels( sampleDT$Dog.person ))` | `r length(levels( sampleDT$Dog.person ))`

The `str()` R function displays a lot of this information all in one call. Useful to look at, as opposed to the above functions that you might want to use programatically:
```{r}
str(sampleDT)
```

Use the built in R function `summary()` on a row to get more details - the function returns different information depending on the data type::
```{r}
summary(sampleDT$Followers)
summary(sampleDT$Dog.person)
```

## Display a quick graph
Or how about a quick graph of the `Followers` column to get a sense of the same data:
```{r}
    par(las=2) # Want the titles to be horizontal
    barplot( sampleDT$Followers, horiz=TRUE, names.arg = sampleDT$Person)
```

I have a separate [cheatsheet on creating a bar chart from MySQL data](quick-barchart-R-cheatsheet.html) and another [cheatsheet on graphing web usage statistics](web-usage-R-cheatsheet.html).

## Find unique values in a column
The `Province` column has a limited number of values - here's a sorted list of the unique values:
```{r}
provinceUniqueSorted <- sort(unique(sampleDT$Province))
cat(provinceUniqueSorted)
```
# Clean up the values in a column
We loaded the `Points` column as a set of characters, with commas as thousands separators and not all people have a value. We'll change the column in-place so that we can continue working with it as number. (I realize we could/should have loaded this column as numeric to start with, like we did with the `Followers` column and just run `sum(sampleDR$Followers, na.rm = TRUE)` = `r sum(sampleDT$Followers, na.rm = TRUE)`.)

Before: `sampleDT$Points` is: `r sampleDT$Points`

We'll run one command to update the column... working from the inside, out, we'll:

1. remove the commas from the string with `gsub()`
1. convert the string to a number with `as.numeric()`
  
```{r}

sampleDT[,6] <- as.numeric(gsub(",", "", sampleDT[,6]), ignore.na=T)

```

After: once updated, the column `sampleDT$Points` is: `r sampleDT$Points`

### Find the sum of the values in a column
We can now find the sum with:

* `sum( sampleDT$Points )`: `r sum(sampleDT$Points)`
* `sum( sampleDT$Followers )`: `r sum(sampleDT$Followers, na.rm=T)`

# References

[^1]: Documentation for read.table() at https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.table.html
[^2]: More information on the version of `R Markdown` I'm using is available [on theRStudio website](http://rmarkdown.rstudio.com/). They describe it as "an authoring format that enables easy creation of dynamic documents, presentations, and reports from R. It combines the core syntax of markdown (an easy-to-write plain text format) with embedded R code chunks that are run so their output can be included in the final document." Here's a useful [R Markdown cheat sheet](http://shiny.rstudio.com/articles/rm-cheatsheet.html) that I've used often.
