---
title: "Quick Bar Chart in R - historic photo exploration"
author: "David Sky"
output: 
    html_document:
        toc: true
        toc_depth: 3
        includes:
            after_body: include/quick-barchart-common-hits.html

---
# Introduction
I've geotagged a wide range of historic Toronto photos, but I have no idea of the breakdown of the dates across the decades. I thought a quick bar chart would give me insight into the data I have. I suspect the 1950's will have the most photographs - lots of construction going on in Toronto, including the new subway, and photography was getting easier all the time. (I authored this [and my other R cheat sheets](index.html) directly in R Markdown[^1] which automatically displays the R code, runs the code, then includes the output in the HTML.)

# Extract data from the MySQL database
I ran a simple MySQL SELECT statment to generate a CSV file with the data I'll need:

    SELECT code, counter, dateyear, datemonth, title, location FROM histPhotos
        WHERE picType = "historic" AND city = "Toronto"


# Load data using the read.table() R function
Once we have the CSV data file from the MySQL SELECT statement, we can load it into a R data table  using `read.table()` and take a look at the first few rows with `head()` .
```{r}
photoDetailsDT <- read.table("historic-photo-details.csv", 
                       header = T, # Our CSV file has column header information in the first line
                       sep=";", # Our CSV file uses a comma to separate each column
                       # col.names = c("dateTime", "neighbourhood") # Reset column names                       
)
head(photoDetailsDT)
```

# Quick look at the data we have loaded
The first column (`code`) corresponds to the original source archive. We can easily list and count the archives with `levels(photoDetailsDT$code)` and `length(levels(photoDetailsDT$code))` which shows us the `r length(levels(photoDetailsDT$code))` sources as: `r levels(photoDetailsDT$code)`

## Add a new column with decade information
But we want to consider the decade for each photo, so we'll add a new column to the data table and use `gsub()` to replace the last digit of the year with a 0 in all cases.

```{r}
photoDetailsDT$decade <- gsub('.{1}$', '0', photoDetailsDT$dateyear)
```
# Prepare the data we need for the graph

To create a bar graph we need two values - the set of decades, and the count for each of those decades.

## Generate decade list
First we'll generate the unique list of decades and assign it to the variable `decadeList`:

```{r}
decadeList <- sort(unique(photoDetailsDT$decade))
cat("List of decades: ", decadeList, "\n")
```

## Generate counts for each decade
Then we'll use the `table()` and `unlist()` functions to generate a list of the counts for each and store that in a `decadeCount` variable:

```{r}
decadeCount <- table(unlist(photoDetailsDT$decade))
cat("Decade count:" , decadeCount, "\n")
```

# Create the graph
We have the two sets of data, so lets create our bar chart!
```{r}
par(las=2) # Want the titles to be horizontal
barplot( decadeCount, 
         horiz=TRUE, 
         names.arg=decadeList,
         main="Historic photo count by decade",
         xlab="number of photos"
         )
```

# Verifying the results
I'm suprised that the most popular decade was 1910 - a good reminder that we should verify these result. It is a simple process to run some SQL COUNT statements against the original data for a few different decades and compare that to our R calculations:

    SELECT COUNT(dateyear) FROM histPhotos 
        WHERE dateyear LIKE '191%' 
            AND picType = "historic"
            AND city = "Toronto"
    
    
This returns a count of 550, and we'll repeat this with a few more decades and create a table to verify that the SQL count and R count are the same!

Decade | SQL LIKE string | SQL count | R count from `decadeCount[]`
-------|-----------------|-----------|--------
1860 | `LIKE '186%'` | 27 | `r decadeCount[6]`
1910 | `LIKE '191%'` | 550 | `r decadeCount[11]`
1970 | `LIKE '197%'` | 57 | `r decadeCount[17]`

# References

[^1]: More information on the version of `R Markdown` I'm using is available [on theRStudio website](http://rmarkdown.rstudio.com/). They describe it as "an authoring format that enables easy creation of dynamic documents, presentations, and reports from R. It combines the core syntax of markdown (an easy-to-write plain text format) with embedded R code chunks that are run so their output can be included in the final document." Here's a useful [R Markdown cheat sheet](http://shiny.rstudio.com/articles/rm-cheatsheet.html) that I've used often.